---
title: "README: HET11 ELISpot Analysis"
author: "Yvonne Dogariu"
date: "`r Sys.Date()`"
output: html_document
---

# Overview

This project contains the analysis pipeline for HET11 ELISpot data, including importing, cleaning, processing, and plotting ELISpot spot forming units (SFUs) from multiple plates and timepoints.

The main goals are to:
- Import ELISpot data from Excel files with multiple sheets
- Annotate and clean the data
- Calculate background-subtracted values
- Generate summary statistics
- Visualize results across timepoints and stimulations

---

# Downloading the Project from GitHub

To download this project, you can clone the GitHub repository using the following command in your terminal or command prompt:

```{bash}
git clone https://github.com/dogariuy/HET.git
```

# Configuration Instructions

Ensure R (version 4.0 or higher recommended) is installed.

Set your working directory to the project root folder after cloning.

Confirm that the directory structure includes:

  data/raw/ for raw Excel input files
  results/figures/ for output plots
  temp/ and processed/ for intermediate and final data files

# Requirements

The analysis requires the following R packages:

- `readxl`
- `ggplot2`
- `dplyr`
- `tidyr`
- `stringr`
- `scales`
- `writexl`
- `viridis`
- `lintr` (optional, for linting)
- `formatR` (optional, for formatting code)
- `docstring` (optional, for function documentation)

You can install them using:
```{r}
install.packages(c("readxl", "ggplot2", "dplyr", "tidyr", "stringr", "scales", "writexl", "viridis"))
# lintr, formatR, and docstring can be installed similarly if desired
```

# Operating Instructions

Place your raw data Excel file (ELISpot data for git.xlsx) inside data/raw/.
Open and run the R Markdown or R script sequentially.
Check temp/ and processed/ folders for saved cleaned data outputs.
Generated plots are saved as PNG files under results/figures/.
Modify experiment IDs or file paths inside the script if analyzing new data.

# File Manifest
  data/raw/ELISpot data for git.xlsx — Raw input Excel file with multiple sheets.
  scripts/ (optional) — Contains the R Markdown or R scripts with analysis code.
  temp/ — Cleaned intermediate Excel outputs.
  processed/ — Final processed datasets.
  results/figures/ — PNG plots visualizing the ELISpot results.
  README.Rmd — This README file.

# Code overview

1. Load Libraries
Loads all necessary R packages.

2. Set Directories
Sets working directories for input, output, and main project folder.

3. Import Excel Files
Defines and uses the function read_plate_data() to read multiple sheets from the Excel file, corresponding to plate layouts and ELISpot data for each experiment ID.

4. Data Wrangling and Processing
Assigns meaningful row and column names to plate layout data.

Reshapes plate data into long format with well IDs.

Merges ELISpot counts with stimulation annotations and timepoints.

Cleans data by removing NAs and converting variables to factors.

Computes average counts per stimulation and plate.

Performs background subtraction using DMSO controls.

Applies a lower limit of detection (LLOD) to replace zero or negative values.

Converts the data into tidy long format for plotting.

5. Save Processed Data
Writes cleaned and processed data to Excel files in appropriate folders (temp/ and processed/).

6. Plotting
Generates boxplots with individual donor points and lines showing ELISpot counts (background-subtracted and log-scaled) across different timepoints and stimulations. Plots are saved as PNG files in the results/figures/ directory.

# How to run the analysis

Place the Excel file "ELISpot data for git.xlsx" in the data/raw/ folder.

Open the R Markdown or script file containing the analysis.

Run all chunks sequentially.

Processed data files will be saved in temp/ and processed/.

Plots will be saved in results/figures/.

# Copyright and Licensing
© 2025 Yvonne Dogariu. All rights reserved.

This project is licensed under the MIT License — see the LICENSE file in the repository for details.

# Contact Information
For questions or assistance, please contact:

Yvonne Dogariu
GitHub: https://github.com/dogariuy

# Known Bugs
The current code assumes a fixed plate layout size (8 rows × 12 columns), but
fewer rows also work. We did not test fewer columns, but it should work.

# Troubleshooting
Ensure the Excel file has sheets named exactly as expected (<experiment_id>_layout and <experiment_id>_data).
Use sessionInfo() in R to help diagnose package/version conflicts.

# Dependencies
## R version 4.5.1
### To install R version 4.5.1: 
Windows: Download the R 4.5.1 installer for Windows from https://cran.r-project.org/bin/windows/base/

After installation, you can check the version of R by running the following in your R console:
```{r}
R.version 
```

## Operating system: Windows
## Package versions
 TempPackage_1.0   
 formatR_1.14      
 lintr_3.2.0       
 viridis_0.6.5     
 viridisLite_0.4.2 
 writexl_1.5.4     
 scales_1.4.0     
 stringr_1.5.1     
 tidyr_1.3.1       
 dplyr_1.1.4       
 ggplot2_3.5.2     
 readxl_1.4.5    
 
 

### To install packages at specified versions:
```{r}
# Install the devtools package (if not already installed)
install.packages("devtools")
library(devtools)

# Install required packages at specific versions
devtools::install_version("formatR", version = "1.14", repos = "http://cran.us.r-project.org")
devtools::install_version("lintr", version = "3.2.0", repos = "http://cran.us.r-project.org")
devtools::install_version("viridis", version = "0.6.5", repos = "http://cran.us.r-project.org")
devtools::install_version("viridisLite", version = "0.4.2", repos = "http://cran.us.r-project.org")
devtools::install_version("writexl", version = "1.5.4", repos = "http://cran.us.r-project.org")
devtools::install_version("scales", version = "1.4.0", repos = "http://cran.us.r-project.org")
devtools::install_version("stringr", version = "1.5.1", repos = "http://cran.us.r-project.org")
devtools::install_version("tidyr", version = "1.3.1", repos = "http://cran.us.r-project.org")
devtools::install_version("dplyr", version = "1.1.4", repos = "http://cran.us.r-project.org")
devtools::install_version("ggplot2", version = "3.5.2", repos = "http://cran.us.r-project.org")
devtools::install_version("readxl", version = "1.4.5", repos = "http://cran.us.r-project.org")

```

## renv version control

Install `renv` (if not already installed):
```{r}
install.packages("renv")
```

Restore the renv locked file-encoded package and R versions automatically by running this code
```{r}
renv::restore()
```
 